How to publish in Biological Sciences: a guide for the uninitiated

### 2021-06-10 4.05-paywall.Rmd
Add files via upload
# UPLOAD working version that prints to pdf and gitbook

(ref:journal-video) Stuffing envelopes with jouranl copies seemed like an endless task. View [on YouTube](https://youtu.be/sJReMVz7Yxk).

@@

At the time, I was aware of Open Access and considered this as an option for the journal. Open Access would have required that someone pay for the type setting, and the society would still have to pay for an online dissemination platform, given that they did not have their own platform. This would have meant that authors paid for getting their manuscripts into print. And then, like today, the decision was that our authors would not be willing to pay. Other, richer, societies were able to go Open Access with the costs being covered by the authors. For some this became an incredibly successful model, with submissions increasing as well as the Impact Factor. They demonstrate that Open Access is possible on an independent platform.

@@

## What will it take to break the vicious cycle?
We need new models for publishing. Society journals are still kings in this game and ultimately hold the cards for moving away from filling the pockets of publishing companies. What we have seen in recent years is that journals can come from nowhere to become dominant players in the system. Think PLoS ONE, and the even more recent Scientific Reports. These mega Open Access journals didn’t exist 10 years ago. And they don’t need to exist 10 years from now. What is needed is for the actual costs of publishing (not that currently inflated by publishers) to be covered by the institutions that employ the academics. This could cover type-setting and the additional IT infrastructure (on-line submission system and online dissemination platform). Most (if not all) societies are not-for-profit organisations, and only need to cover the costs of publishing.

@@

The solution then would be for our learned academic societies to come to agreements directly with university libraries. Today, publishing is less about type-setting (which many top publishers outsource and do very poorly), and more about dissemination. This is something that our library and academic librarians have been doing for decades, and do far better and far more cheaply than any publishing house. Given the choice, most of us would prefer to entrust our academic endeavors to our own libraries rather than to for-profit publishing houses.

There are more reasons why it makes sense for libraries to take on the roll as publishers. Most of us are employed by universities or research institutions that also fund our libraries. Linking the work we do (writing, reviewing and editing) more closely with our institutions would result in a greater appreciation for this part of our workload. Editors and associate editors will appreciate that they get little credit from their institutions for a considerable extra amount of work that they perform.

Libraries have fantastic networks, and are our professional long-term storage partners. They developed efficient and impressive information technology (IT) long before it hit most academic departments. Their inter-library networks are what we now need to disseminate the knowledge that we generate without any walls.

The idea of libraries as the new publishers isn't new. Raju and Pietersen [@raju2017library] proposed this as a solution in Africa. Here I extend the same idea as an exclusive way of publishing academic journals for the world.

## We need to give up our addiction to fancy layouts
Once the storage and dissemination of our contributions are taken care of, the only service left from the publishers is a fancy layout. This is mostly a historical legacy which I've talked more about in the past (see above). I have to admit that I really like seeing my work being nicely produced and printed. But I’m happy to give this up if it means demolishing paywalls. In reality, LaTex can solve most of these problems so that we simply use the journal (library) produced template, that will need minimal manipulation afterwards.

We also need to persuade our employers to stop their obsession with publishing metrics, for example by adopting the [San Francisco Declaration on Research Assessment](https://sfdora.org/read/) (known as DORA).

No doubt, there will be some institutions that will invest extra to have nicer layouts. But I feel confident that this will not change the impact factor, or any other metrics, as academics will value the content for what it contains rather than what it looks like. Admittedly, nothing about the contents of the highest ranking journals suggests that impact factor is consistently related to research quality.

@@

>>>>

(ref:journal-video) Stuffing envelopes with journal copies seemed like an endless task. View [on YouTube](https://youtu.be/sJReMVz7Yxk).

@@

At the time, I was aware of Open Access and considered this as an option for the journal. Open Access would have required that someone pay for the type setting, and the society would still have to pay for an online dissemination platform, given that they did not have their own platform. This would have meant that authors paid for getting their manuscripts into print. And then, like today, the decision was that our authors would not be willing to pay. Other, richer, societies were able to go Open Access with the costs being covered by the authors. For some this became an incredibly successful model, with submissions increasing as well as the [Impact Factor](impactfactor.html). They demonstrate that Open Access is possible on an independent platform.

@@

## How do publishers make their money? {#money}

In the original publishing model, scholarly societies, such as The Royal Society, paid publishers to type set and print their publications. These in turn were bought by the society for their members and any libraries who subscribed to the society. As institutions and their libraries grew internationally, so too did the subscription base such that there was a profit to be made from the cost of the subscriptions for the paper content. Even back in these early Victorian days of scholarly publishing, publishers to started their own journals (aside from any society) as they recognised their superiority in distribution [@brock1980development]. This model continued over many decades with some publishing houses growing as they acquired more journal titles from societies because there was a lot of profit to be made. The peculiarity of scholarly publishing includes the unpaid nature of authors, editors and reviewers (all of the content), and the non-equivalence of the output: that is to say that just because you have access to articles in _Nature_, they are not equivalent to those being published in _Science_ - as a scientist you need to read articles from both of these publications.

Interestingly, the dawn of the internet was predicted as the downfall of the academic publisher Elsevier, because it promised the ability for academics to share their content for free [@lariviere2015big]. Despite this Elsevier's profit in 2012-13 was USD 2 billion, with a profit margin of ~40% [@lariviere2015big]. This profit margin is so large that some have compared it favourably with street gangs selling drugs [@buranyi2017staggeringly]. In the Biological Sciences, Elsevier, Springer-Nature, Wiley and Taylor-Francis control around 50% of all papers published [@lariviere2015big]. Selling to university libraries was no longer made on print copies, but on packages of journals from each of these big publishers. The deals made by individual libraries were (and are) made in strict confidence, the exact sums are unknown, but the deals made fall into the hundred and thousands of USD for a set of publications that cover (for example) the science content of one of these publishing houses. The penalty for not subscribing is that your researchers reach a [paywall](paywall.html) when they try to access a particular paper. You will find, as I have, the luxury of visiting a very wealthy institution and having access to just about every journal title you can think of. If you are not at one of these elite institutions, then you will face a demand for money when you try to access a title that you do not subscribe to.

The cost of accessing a single article comes at between 25 and 50 USD [@hagve2020money]. For me, and a great deal of other researchers, there are other ways to [get around the paywall](http://www.howtowriteaphd.org/paywall.html). But for some people, especially members of the public who simply want to access the information, they may simply pay the price. Imagine then the potential profit to be made from a society that has a large back-catalogue of content. There has been a scramble to sign up editors from society journals with tens of thousands of old journal articles at their command. The publisher offers to scan the entire back-catalogue and then offer it on sale. The society gets a paltry 10%, while the publisher fishes with tens of millions of articles. Even if they only sell 1% of these in 1 year, that makes their profit in the millions of USD.

In 2014, universities in Germany abandoned all subscriptions of journals by the Dutch publishing house Elsevier, as they couldn't keep up with the 30% price increase over the previous 5 years, making the average cost ~$4700 per journal per year [@vogel2014german]. Despite this, Elsevier continued to allow these universities access to their content for free, before negotiating another (secret) more amenable deal to them. Somehow the publishers had managed to turn the ease of distribution of articles over the internet to their advantage, via effective paywalls. But this was not the last of their master strokes.

[Open Access](openaccess.html) titles came about in the early 2000s as the need for hard printed copies fell away as academics became comfortable handling all of their publications as pdfs. In the Biological Sciences, PLOS ONE was the title that revolutionised the market, quickly becoming the single biggest journal. What caught the eyes of the publishers was the article processing charge (APC) that academics seemed to be prepared to pay. When they first started, PLOS ONE cost ~1300 USD for its APC [@khoo2019article]. Publishers quickly started titles with similar [no impact factor](#noif) goals. Competitive titles began at the same cost as PLOS ONE, but once these titles became established, with higher Impact Factors, they quickly increased their APC [@khoo2019article]. Commercial publications (as oposed to non-profit journals) have become adept at pricing their titles toward what libraries and authors are prepared to pay, such that commercial publications are regularly more expensive than non-profit journals [@bjork2015article]. In their analyses of drivers of APC for OA journals, Budzinski et al [-@budzinski2020drivers] found several variables that explained the vast difference in APCs of different journals. Impact Factor played an important role, but so too did market power of the publishers (presumably the ability to distribute the title and content), the hybrid model (whether it is gold OA, or ******), and the concentration of disciplines (*****)

Librarians analysed content that was [Open Access](openaccess.html) finding that it was more highly cited, and so academics began to look for the potential to publish [Open Access](openaccess.html). The publishers were happy to oblige, and in the early days conducted 'double dipping' charging both the APC for the authors, and collecting full revenue for subscription to hybrid journals by universities [@pinfield2016total]. It was only in May 2018 that [Knowledge Unlatched](https://knowledgeunlatched.org/2018/05/knowledge-unlatched-announces-its-fifth-pledging-round-ku-select-2018-2/) formed an alliance to prevent double-dipping in Open Access publications.

The end to double-dipping, as mandated by governments [@pinfield2016total], has led to even more complicated publishing deals known as **transformative publisher agreements**. With these agreements, researchers from universities that have full subscriptions to certain publishers will receive free Gold OA for any journals in the package. In effect, the price of the OA is offset against the cost of the subscription. As you may have noticed, this will quickly start a bias in the publishing trends of academics from some (middle income) institutions publishing with certain titles, because they receive the benefit of OA. Those from privileged institutions will continue to publish anywhere they can, and the real losers will be those from developing, especially middle-income countries, who cannot afford transformative agreements, and cannot afford to provide their researchers with costs to cover APCs [@measey2018europe]. For these researchers, there will be a reduced set of journals that they can subscribe to. This is the current reality that many of us find ourselves in.

On the face of it, Plan S makes total sense [@schiltz2018science]. If all papers published by European researchers are Open Access, then everyone can read them and there is no more need for paywalls. Unfortunately, the way in which Plan S has been implemented has only been for wealthy European institutions. Others are left to continue to pay a hefty APC price for OA publishing, and their libraries pay a subscription to access journal content [@hagve2020money].

Another way that publishers make money is through [copyright](#copyright) on the content of their very large back catalogues. There are a lot of text books written for undergraduate students, and this is a large profit making publishing model in and of itself. When you look at the examples given in text books, they are often taken directly from peer reviewed manuscripts. The publishers of the text book will pay for the content that they get from the pages of journals where they are under copyright. This can be very lucrative for the publishers who retain copyright from the authors. Of course, the authors get nothing even though it was they that created the content. The governments that funded the authors also get nothing. The advent of Creative Commons licencing has meant that it is now very easy to make the content of journals available for anyone to use. Happily, most Open Access journals have taken this route.

### How could such a profitable business lead to making a loss?
To be fair, we should also consider that some publishers have not (always) been able to turn a profit, even when their turnover is in millions of USD. One such beleaguered publisher is PLOS. Despite growing exponentially in the early noughties, to become the world's biggest academic publisher in 2013, PLOS has not managed to turn a profit since 2015 [@davies2019plos]. At the time of writing (May 2021), the APC on a PLOS ONE article was USD 1749, while other PLOS titles ranged from USD 2100 to 4000 (see [PLOS fees site](https://plos.org/publish/fees/)). This begs the question, how much does it cost to process an article? Data from the PLOS Financial Overview suggests that processing your article costs USD ~315, with another USD ~262 to keep it online.

So where does the rest of the money go? PLOS spend the bulk of your APCs on the editorial aspects of publishing. Remember that PLOS don't pay academic editors, but they do have huge financial burdens of the salaries of their own editorial staff. PLOS is based in San Francisco, California, which just happens to be one of the most expensive cities in the world to live in. Chief among their many options are to raise the costs of your APCs in order to pay their staff, or move their operation to somewhere cheaper. Interestingly, the other two large OA outfits, MDPI and Frontiers, are both based in Switzerland, possibly one of the most expensive countries in the world in which to pay salaried staff.

Most recently, PLOS announced a new funding model to keep themselves afloat that spreads the cost of their APC among the authors, and aims to cap profits at 10% [@else2021open-access]. The new scheme, Community Action Publishing, sees a return to the institutional subscription model where institutions will be charged proportionately at the rate their academics publish in PLOS. Great if your institutions can afford to opt in. If not, you can expect to face your proportion of the APC now as a co-author (instead of only the corresponding author paying).

Time will tell whether PLOS can make their new financial model work in order to keep their staff in the comfort to which they have become accustomed. No doubt we will see more of these new funding models as the increasingly bloated OA market attempts to consolidate itself, while still feeding on the funds of the global tax payer.

## Making a profit - In summary**

 - From society journals
   + old subscription model (mostly obsolete)
   + selling bundles to university libraries (undisclosed sums - millions of USD)
   + selling articles of large back catalogues (~10% goes to society)
 - From their own journals
   + old subscription model  (mostly obsolete)
   + selling bundles to university libraries (undisclosed sums - millions of USD)
   + selling articles of large back catalogues
 - From [Open Access](openaccess.html)
   + APC
   + Double dipping
   + Transformative publisher agreements
 - From [copyright](#copyright)
   + Any content that anyone wants to use (such as in a text book)

In case you hadn't realised, there is no advertising (at least in Biological Sciences - some medical journals do have advertising). The only time you see adverts is when you are browsing the content of Nature and Science. Oddly, this doesn't make them any cheaper to publish in, but presumably does increase the profits for these particular titles.

### What about academic books?
The story about the move from print to eBooks in academic books shows some even more remarkable disparities in the prices of eBooks compared to their physically printed paper counterparts (see Figure \@ref(fig:eBook-prices)). University libraries are required to buy versions that are specifically licensed for university use. The copyright law means that universities are not allowed to scan their paper copies and make these available. But in order to buy the university library copy, the price paid might be as much as 10 times that of the paper edition (Figure \@ref(fig:eBook-prices)).

(ref:eBook-prices) **The cost of eBooks compared to printed copies of the same books.** In crowdsourced data compiled by people in the Campaign to investigate the academic eBook market, you can see that most of the titles have a much more expensive eBook version. Note that in all cases the eBook licence is the equivalent of a single printed book: i.e. a 1 user copy. Data as entered on 17 May 2021.

```{r eBook-prices, echo=FALSE, fig.cap="(ref:eBook-prices)"}
if(knitr::is_html_output(excludes="markdown")) knitr::include_url("https://johnmeasey.github.io/eBooks-prices/index.html", height="450px") else knitr::include_graphics("figures/eBook-prices/eBook-prices.png")
```

More than this, publishers appear to inflate the prices of their eBooks without warning, but as much as 200% [@anderson2021campaign]. In addition, there are moves to make licensing annual for universities, with examples where once-off purchases have become unavailable as the publisher has moved them to an annual pricing model. While copyright law would protect you as an individual from any publisher that was undertaking such terrible behaviour, UK universities appear to be without recourse, such that they are now petitioning the Minister for Higher Education [@anderson2021campaign].

## What will it take to break the vicious cycle?
We need new models for publishing. Society journals are still kings in this game and ultimately hold the cards for moving away from filling the pockets of publishing companies. What we have seen in recent years is that journals can come from nowhere to become dominant players in the system. Think PLoS ONE, and the even more recent Scientific Reports. These mega Open Access journals didn’t exist 10 years ago. And they don’t need to exist 10 years from now. What is needed is for the actual costs of publishing (not that currently inflated by publishers) to be covered by the institutions that employ the academics. This could cover type-setting (fulfilling our irrational desire for '[fancy layouts](#layout)' and the additional IT infrastructure (on-line submission system and online dissemination platform). Most (if not all) societies are not-for-profit organisations, and only need to cover the costs of publishing.

### A role for scolarly societies and libraries {#societies}
Scholarly societies have an important role to play in leading change in the scientific publishing model. 
(#change)  make the shift from [closed to open science](#openscience) [overlay journals](#overlay) [Diamond Access](#diamond) libraries(#money)
[@potvin2018diversity]
**Section to write**

@@

Like others, I think that the logical solution to our problem with publishers is to turn instead to our university libraries to curate our academic outputs [@fang2012reforming]. There are clear reasons why it makes sense for libraries to take on the roll as publishers. Most of us are employed by universities or research institutions that also fund libraries. Linking the work we do (writing, reviewing and editing) more closely with our institutions would result in a greater appreciation for this part of our workload. Libraries have fantastic networks, and are our professional long-term storage partners. They developed efficient and impressive information technology (IT) long before it hit most academic departments. Their inter-library networks are what we now need to disseminate the knowledge that we generate without any walls. The idea of libraries as the new publishers isn't new. Raju and Pietersen [@raju2017library] proposed this as a solution in Africa. Here I extend the same idea as an exclusive way of publishing academic journals for the world.

### Moving back to publishing through scholarly societies
without publishers we still need to organise academics at a level beyond the individual and spanning institutions. This means elevating the importance of scholarly societies [@harington2020importance], with all of the additional benefits that these bring (e.g. networking, conference organisation, newsletters, socialising, mentoring, stewardship, community of practice, ethics, outreach, etc.). However, far from relying on the publishers be the sources of income for societies [@harington2020importance], we need a professional society model (such as is seen in medicine and other vocationally orientated disciplines) whereby the relationship between academics and their societies is strengthened through their institutions.

## We need to give up our addiction to fancy layouts {#layout}
Once the storage and dissemination of our contributions are taken care of, the only service left from the publishers is a fancy layout. This is mostly a historical legacy (see above). I have to admit that I really like seeing my work being nicely produced and printed. But I’m happy to give this up if it means demolishing paywalls. In reality, LaTex can solve most of these problems so that we simply use the journal (library) produced template, that will need minimal manipulation afterwards. I feel sure that those who are hung up on the importance of their layout can find undergraduates at their own institutions who will be happy to provide layout services for a reasonable fee. No doubt, there will be some institutions that will invest extra to have nicer layouts. But I feel confident that this will not change the impact factor, or any other metrics, as academics will value the content for what it contains rather than what it looks like. Admittedly, nothing about the contents of the highest ranking journals suggests that impact factor is consistently related to research quality.

If our futures lie with the [overlay](#overlay) model of publishing, then I suggest that we need to have less reliance on international storage of (what are currently) 'preprints', and instead a closer relationship between university libraries and scholarly societies.

@@

## Giving up the obsession with metrics
Another key move will be having all academic institutions and funders adopting the [San Francisco Declaration on Research Assessment](https://sfdora.org/read/) (known as [DORA](#DORA)) in order to stop hiring staff based on their publishing metrics, and the negative impacts that this is known to have [@casadevall2012reforming]. This will have an impact on many aspects of science including the current way in which science is funded on a '[winner takes all](#winnertakesall)' basis [@casadevall2012reforming]. Many suggestions for reforming science are out there, but when the participants are still ruled by the cult of metrics, it will be hard to see the necessary reform taking place [@fang2012reforming].
